Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-04-27 paper/main.tex
some reordering, more headlines

\begin{methods}
\section{Methods}
The foundation of our work are 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant genomes of the JGI Phytosome 13 database \citep{goodstein2012phytozome}. Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome group is made up of 19 and 6 genomes, respectively, and was chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread. We then experimented which genomes to use for during training from the pool of non-test genomes while evaluating the performance of models on the remaining ones. We call the genomes that were ultimatily not used during training the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance of the combined set of validation and test genomes as the validation genomes are by far the largest set. We also did not find a noticible difference in performance between those two groups.

We pre-processed and stored the genomic information by using Geenuff \citep{denton2019}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. The encoding of the genomic sequence is done in line with the IUPAC nucleic acid notation and the functional annotation used as labels during training is transformed to a one hot encoding with the four classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table \ref{tab:arrays} for a more detailed description of the generated data formats.

The classes are not balanced equally and the tradeoff between the number of intergenic and intronic bases is quite different in both groups of genomes. Table \ref{tab:statistics_all} shows the ratio of class occurences, together with other overall metrics.

@@

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation. As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to blank out those bases during training by using the sample weights described in Table \ref{tab:arrays}. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes only can be found in Figure S6.

For the training, we divided each continous genomic sequence into 20.000 bases long subsequences, which are then used as input or label for the neural network. We appended zero padding if the subsequences are not long enough. If a subsequence is fully marked as erroneous, we do not include it anywhere.

The selection of training genomes requires balancing multiple conflicting tradeoffs. On the one hand, we do want to train with as much data as possible, but not all data has a good enough quality to enable a powerful generalization. It is also very desirable to have a broad spread in phylogenetic distance as well as genome size and average gene length as the model is increasingly unlikely to generalize well beyond the borders given by the training data. However, if the genomic patterns inside the data are too different from each other, it may be difficult for the model to generalize as well. It was also important to us to get experimental results within a couple of days as we tested many different data and model configurations, which limited the overall size of the training set and the model.

@@

\subsection{Model Architecture}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization \citep{ba2016layer} between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. The hyperparameters were optimized by a combination of manual and automatic optimization. Automatic optimization was carried out by using either the TPE algorithm \citep{bergstra2011algorithms}, random search or grid search depending on the situation. We used NNI \citep{nni2019} to facilitate the search. All relevant hyperparameters can be found in Table \ref{tab:params} or in the Helixer source code repository.

>>>>

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant genomes of the JGI Phytosome 13 database \citep{goodstein2012phytozome}. Both data groups were used seperately during training as well as during evaluation.

At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome groups are made up of 19 and 6 genomes, respectively, and were chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread.

The selection of training genomes also requires balancing multiple conflicting tradeoffs. On the one hand, we do want to train with as much data as possible, but not all data has a good enough quality to enable a powerful generalization. It is also very desirable to have a broad spread phylogenetically as well as in genome size and average gene length as the model is increasingly unlikely to generalize well beyond the borders given by the training data. However, if the genomic patterns inside the data are too different from each other, it may be difficult for the model to generalize, too. It was also important for us to get experimental results within a couple of days as we tested many different data and model configurations, which limited the overall size of the training set and the model.

We call the genomes that were ultimatily not used during training from the original split the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance on the combined set of validation and test genomes as the validation genomes are by far the largest set. We also did not find a noticible difference in performance between those two groups.

\subsection{Data Pre-processing}
We first pre-process and store the raw genomic information by using GeenuFF \citep{denton2019}. GeenuFF is a tool for checking and exploring genomic data, that stores all information inside a SQL database. Training and evaluation-ready data is generated by querying this database and then transforming the returned data into a numerical format suitable for machine learning. The encoding of the genomic sequence is done in line with the IUPAC nucleic acid notation and the functional annotation used as labels during training is transformed to a one hot encoding with the four classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table \ref{tab:arrays} for a more detailed description of the generated data types.

@@

During data generation, we query for the longest transcript of each gene and disregard FASTA sequences that have no functional annotation. GeenuFF checks the genomic annotations for potential errors during import and is able to mark those areas. We use this information to effectively blank out those bases during training by using the sample weights described in Table \ref{tab:arrays}. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and GeenuFF marks a potentially large intergenic region for it. Table \ref{tab:statistics_all} shows statistics about the masking.

For the training itself, we divide each continous genomic sequence into 20.000 bases long subsequences, which are then used as input or label for the neural network together with the sample weights. We append zero padding if the subsequences are not long enough. If a subsequence is fully marked as erroneous, we do not include it anywhere.

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization \citep{ba2016layer} between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. We also use class weigths as the classes of the input data are not balanced equally enough, at least not for our network, and does also vary greatly between animals and plants. For example, the tradeoff between the number of intergenic and intronic bases is quite different in both groups (see Table \ref{tab:statistics_all}).

The hyperparameters were optimized by a combination of manual and automatic optimization. Automatic optimization was carried out by using either the TPE algorithm \citep{bergstra2011algorithms}, random search or grid search depending on the situation. We used NNI \citep{nni2019} to facilitate the search. All relevant hyperparameters can be found in Table S3 and S4 or in the Helixer source code repository.
